Physiological sensor based mental workload assessment: A multimodal deep learning approach

### 2021-05-13 0.manuscript/manuscript.tex
Final version thesis

\title{Physiological sensor based prediction of mental workload: A multimodal deep learning approach}

@@

\begin{abstract}
The assessment of mental workload by means of physiological sensors has gained in popularity in recent years. We utilized three physiological sensors from which mental workload was predicted, being electroencephalography, photoplethysmography and galvanic skin response. Deep learning poses a promising, yet novel, approach towards the modeling of such sensory data. Four deep neural network architectures were created and contrasted in their performance. Three out of four were specified to predict workload based on on a single of the aforementioned physiological sensors, whilst the fourth architecture was a joint neural network combining all physiological signals into a single model. In this study we aimed to identify 1) which physiological sensor constitutes the most adequate predictor of mental workload, and 2) whether a joint approach utilizing multiple sensors is preferable over the approaches utilizing just one. Altogether the electroencephalography-only architecture was found to perform best, closely followed by the galvanic skin response-only architecture,  both displaying a mean absolute error of 0.110 and 0.119 (scaled on 0-1) respectively. Against expectations, the joint approach was outperformed by both aforementioned architectures, displaying a mean absolute error of 0.128 (equally so scaled on 0-1). The photoplethysmography-only architecture was found to perform worst, displaying a mean absolute error of 0.136. In the discussion section we presented several explanations that help rationalize this anomaly, one being due to an imperfect model architecture for the joint neural network, and another being due to overfitting resulting from a limited amount of data. Regardless of the unexpected character of this result, we still recognize that differences in performance were of rather modest size: all four neural network architectures performed fairly adequately. 
\end{abstract}

@@

\begin{large}
{\setlength{\parindent}{0cm}
Methodology and Statistics for the Behavioural, Biomedical and Social Sciences
}
\newline Utrecht University, the Netherlands
\vspace{18mm}
\newline MSc Thesis Bart-Jan Boverhof (6000142)
\newline Physiological sensor based prediction of mental workload: A multimodal deep learning approach
\newline May 2021
\vspace{18mm}
\newline Supervisor:
\newline Prof. Dr. Ir. Bernard Veldkamp
\vspace{10mm}
\newline Second grader:
\newline Prof. Dr. Ir René Eijkemans
\vspace{10mm}
\newline Preferred journal of publication: Frontiers in Neuroscience
\end{large}
\end{titlepage}

@@

A commonly employed method for the assessment of workload in an experimental setting is the well established NASA-Task Load Index questionnaire, hereafter referred to as \enquote{NASA-TLX}. This method embodies several dimensions, jointly inquiring into the amount of perceived workload \cite{hart2006nasa}. However, due to the ever-increasing availability in computing power characterizing recent times, a novel approach to workload assessment has gained in popularity. This promising approach constitutes the use of physiological sensors, worn throughout the entire duration of the experiment. An approach like this poses the advantage that the development of perceived workload may be monitored throughout the entire experiment, enabling the possibility to precisely monitor the impact of a specific task on a participant's level of workload. Another advantage is that the method may be personalized, which could simply be done by creating and training an individual model for each participant. By doing this, the person-specific character that is inherent to workload is acknowledged, the importance of which was stressed in the opening paragraph of this paper.

With the current research we explore the feasibility of an approach towards workload assessment by means of various physiological sensors, hereafter referred to as \enquote{modalities}. The three utilized modalities are electroencephalography (EEG), photoplethysmography (PPG) and galvanic skin response (GSR). EEG is a technique that monitors electrical activity in the brain, and is often utilized in the assessment of workload \cite{berka2005evaluation, craik2019deep}. EEG is found to be amongst the most adequately performing techniques for this task \cite{hogervorst2014combining}. GSR is a technique that measures electrical conductance of the skin, and equally so is a widely adopted method for workload assessment \cite{nourbakhsh2012using, zhou2015dynamic}. Lastly, also heart-rate is a widely utilized indicator of workload, mostly obtained through PPG \cite{zhang2018evaluating,jimenez2018using}.

In this project we opted for a deep learning approach towards workload prediction. A range of four different deep neural network architectures, hereafter referred to as \enquote{DNN's} or simply \enquote{networks},  are constructed and contrasted in their performance. Among these are three unimodal DNN's, i.e. networks each utilizing only a single modality. Thus,  one network solely utilizes the EEG modality,  one solely utilizes the GSR modality and one solely utilizes the PPG modality. The fourth is a multimodal network, i.e. combining all three aforementioned modalities into a single architecture. The main advantage of taking a deep learning approach specifically resides in its flexibility. Fitting a model that combines multiple modalities can be done in different, relatively straightforward, ways when taking a deep learning approach \cite{ramachandram2017deep}. We expect the multimodal approach to results in better performance as compared with the unimodal approaches, for predictions are based on richer composite of information. This expectation is indeed in line with previous research, which reported a multimodal approach to perform superiorly \cite{dolmans2020perceived, han2020classification,rastgoo2019automatic, yin2017recognition}.

With this research we 1) seek to explore which individual modality constitutes the most adequate predictor of workload when taking a deep learning approach, and 2) whether a multimodal network architecture is preferable over the three simpler (but much less computationally demanding) unimodal architectures. This comparison will mainly be propelled by contrasting network performance, however when drawing conclusions, considerations regarding computational costs will also be taken into consideration. With this research we aim to deepen the knowledge of taking a deep learning approach towards workload prediction with physiological sensors in an experimental setting, which as of yet, is still a novel and fairly understudied topic.

@@

\subsubsection*{EEG Network Architecture}
An overview of the complete literature EEG applications with deep learning was presented by \citeA{craik2019deep}, who reported 16\% of all publications to constitute workload assessment. The lion's share of these publications utilized either a deep belief network or convolutional neural network (ConvNet). \citeA{tabar2016novel} proposed a ConvNet and stacked auto-encoder (SAE) hybrid approach towards workload classification with EEG. Their proposed network constituted two different parts. Firstly, raw EEG data was fed into a single convolutional layer, with the objective of extracting features. Secondly, the the extracted features fed into the SAE part of the network,  which encompassed a stack of dense-layers designed to ultimately result in a classification of workload.  An accuracy of 90\% was acquired with this architecture \cite{tabar2016novel}. A similar approach was proposed by \citeA{schirrmeister2017deep}, who contrasted different DNN architectures in their ability to adequately decode EEG signals. Several ConvNet architectures were contrasted against the baseline method for EEG decoding, filter bank common spatial pattern (FBCSP). A deep ConvNet, a shallow ConvNet, a deep-shallow hybrid ConvNet and a residual ConvNet were contrasted with the FBCSP baseline method. Both the deep and shallow ConvNets were found to reach at least similar, and in some regards better classification results as compared with the FBCSP baseline approach. Altogether, a deep ConvNet with four convolutional-max-pooling blocks was found to perform best, exhibiting a classification accuracy of 92.4\% \cite{schirrmeister2017deep}.

\subsubsection*{GSR Network Architecture}
\citeA{sun2019hybrid} explored various deep-learning approaches towards the classification of emotional states with GSR. Despite that this application differs from workload prediction, inspiration may still be drawn from their utilized architecture. Various models were explored, amongst others a support vector machine, a ConvNet and a long-short-term-memory (LSTM) network. In addition to these, a hybrid ConvNet-LSTM approach was explored as well. This hybrid approach constituted four convolutional blocks followed by a single LSTM layer, and was found to perform best with an accuracy of 74\% \cite{sun2019hybrid}. \citeA{dolmans2020perceived} contrasted several physiological signals in their ability to predict workload, each of which by taking a deep learning approach. One of these modalities constituted GSR, for which a ConvNet-LSTM hybrid network architecture was utilized to predict workload. The performance of this DNN was contrasted with a network consisting solely of fully connected dense layers. Conform with the previously mentioned findings by  \citeA{sun2019hybrid}, the ConvNet-LSTM hybrid architecture was found to perform best, displaying an absolute difference between predicted and true label of 0.180 (scaled on 0-1). This network architecture incorporated two convolutional blocks, followed by two LSTM layers \cite{dolmans2020perceived}.

@@

\subsubsection*{Multimodal Fusion Strategy}  
When creating a DNN that combines multiple modalities into a single network,  the information streams that stem from the different subparts of the network (each corresponding to one of the modalities) are required to be combined, i.e. \enquote{fused}, at a given point in the network. This is necessary in order to ultimately end up with a single prediction of workload. Fusion can be done conforming several strategies, out of which three strategies as proposed by \citeA{ramachandram2017deep} are considered.

Firstly, early fusion constitutes an approach that fuses data sources before being fed into the network. Early fusing usually proves to be quite challenging, residing in the fact that data streams stemming from different modalities often differ in their dimensionality and sampling rate. In addition, when taking an early fusion approach, the assumption of conditional independence is made implicitly. This assumption is oftentimes unrealistic in practice, for data stemming from different modalities are expected to be correlated \cite{ramachandram2017deep}. Secondly, late fusion refers to the process of aggregating the decisions of multiple separate networks, each constructed for every modality separately. When the data sources stemming from the various modalities happen to be either correlated or different in their dimensionality, late fusion is oftentimes a more feasible approach as opposed to early fusion \cite{ramachandram2017deep}. Lastly, intermediate fusion is the most widely employed fusion strategy for multimodal networks. Data streams are usually fused by a concatenation layer, joining the outputs of the separately defined network parts of each modality. This results in a single joint deep neural network. Several higher-order layers are usually defined in between the concatenation layer and the ultimate prediction layer. The depth of the fusion, i.e. the specified number of higher-order layers, may be chosen by the researcher, posing intermediate fusion to be the most flexible and therefore the most widely adopted fusion strategy \cite{ramachandram2017deep}.

@@

\subsubsection*{Auxiliary Layers in a DNN Architecture}
At the core of a DNN are the aforementioned types of layers, such as convolutional, LSTM and dense layers. Such layers are often accompanied with auxiliary layers in order to ensure network stability. The first type of layers often used in network architectures are batch normalization layers, which have the objective improving network stability. It is beneficial to incorporate a batch normalization layer subsequent to a convolutional layer, but before feeding into the activation layer \cite{ioffe2015batch}.  For a deep dive into batch normalization, we refer the reader to \citeA{ioffe2015batch}. All previously considered network architectures made use of batch normalization layers \cite{biswas2019cornet, dolmans2020perceived, schirrmeister2017deep, sun2019hybrid, tabar2016novel}.

Pooling layers are commonly employed in ConvNets, usually succeeding a convolutional layer with the purpose of reducing dimensionality. The objective of such layers are to down-sample features into a more compact space, hereby only retaining essential information, and thus omitting redundant information. For a more extensive elaboration on pooling we refer the reader to \citeA{lecun2015deep}. All previously considered network architectures utilize max-pooling layers, usually specified subsequent to the activation layer \cite{biswas2019cornet, dolmans2020perceived, schirrmeister2017deep, sun2019hybrid, tabar2016novel}.

The incorporation of dropout layers was initially proposed by \citeA{srivastava2014dropout}. Dropout layers drop a randomly-selected number of neurons and their connections throughout the training process. More specifically, within each full training run through the data, i.e. each \enquote{epoch}, a different randomly selected set of neurons are fixed to be inactive. Doing so helps to reduce overfitting, particularly for complex network architectures \cite{srivastava2014dropout}.  Most of the previously considered network architectures make use of dropout layers \cite{biswas2019cornet, dolmans2020perceived, schirrmeister2017deep, sun2019hybrid}.

\subsubsection*{Hyperparameter Optimization}
Hyperparameter optimization (HPO) is a technique that optimizes network hyperparameters, such as for example learning rate and dropout rate. Substantial advancements in DNN performance have been attained by utilizing HPO, especially in the case of ConvNets \cite{bergstra2012random}. In theory, one could optimize the entire DNN architecture, including the number of neurons/filters in dense/convolutional layers, the number of layers in general, whether to use certain layers, etc. There is, however, a strong relationship between the amount of parameters to be optimized and the computational resources required to do so: optimizing many hyperparameters will inflate computational costs substantially. The Optuna toolbox provides a method for creating a parameter search space, from which values for the hyperparameters may be sampled, and optimization can be performed \cite{akiba2019optuna}.

@@

\subsubsection*{Experimental Setup \& Data Labeling Strategy}
The experimental setting in which our data was collected is the open-source spaceship video-game Empty Epsilon, in which partakers carried out tasks on a virtual spaceship bridge \cite{daid2016empty}. This experiment was instituted by the Brain Computer Interfaces Testbed \footnote{https://bmslab.utwente.nl/} hosted by the University of Twente and carried out in cooperation with Thales Group\footnote{https://www.thalesgroup.com/en/countries/europe/netherlands} Hengelo, the Netherlands. The experiment was divided into a total of 15 tasks, differing in difficulty and hence designed to evoke varying degrees of workload. The total duration of the experiment was roughly three hours.

Respondents were asked to fill out the NASA-TLX questionnaire after each of the fifteen task, inquiring into the amount workload perceived during the respective task. Each questionnaire consisted out of six items. The mean of a subset of four of these items were utilized for labeling of the data. Two items were not incorporated into the scale, for they inquired into the perceived physical demands. Since the experiment took place on a computer whilst participants sat behind a desk, we chose to not incorporate such items. Each item, and hence the newly constructed scale, ranges from 0-20, wherein 0 reflects the lowest possible level workload and 20 reflects the highest possible level workload.  In order to encourage numerical stability, label scores were normalized to reside in-between 0-1.

@@

\subsubsection*{Data Preprocessing \& Partitioning}
For each participant, the EEG, PPG and GSR data streams were cut into windows of eight seconds, resulting in an average of roughly 351 windows per respondent. 80\% of all windows have been used for training, 10\% for validation within the training process and 10\% for the testing of network performance. The partitioning of the windows into training, validation and testing sets was done systematically by selecting every \textit{n-th} window, such that windows over the entire duration of the experiment were represented equally in each partition.  Exactly the same partitions were utilized in the training, validation and testing procedure across the four architectures to ensure a fair comparison.

\subsection{Network Architectures \& Training} \label{section:DNN}
Individuals tend to respond differently towards high workload situations: for some workload manifests itself mostly through an increase in heart rate detected through PPG, whereas for others it manifests as an increase in skin conductivity detected through GSR. To account for such between-person differences, we trained an individual DNN for each participant. Since a total of four architectures are contrasted for 27 respondents, this resulted in a total of 108 trained networks. Equally so, HPO was conducted for each person individually, resulting in 108 sets of optimized hyperparameters. All networks were constructed in Python, by utilizing the deep-learning toolbox PyTorch \cite{paszke2017automatic}.

\subsubsection*{Unimodal Network Architectures}
All unimodal architectures drew inspiration from aforementioned research \cite{biswas2019cornet, dolmans2020perceived, schirrmeister2017deep, sun2019hybrid}. We opted for a ConvNet architecture for all three unimodal network architectures.

The three unimodal DNN architectures are depicted alongside one another in Figure \ref{fig:singlearchitecture}. Each of the unimodal networks may be distinguished by two separate parts, being the convolutional part and the prediction part.  The objective of the convolutional part is to extract features from the data, and consists of four convolutional blocks, each constituting a convolutional layer, a batch normalization layer, the Exponential Linear Unit (ELU) activation function, and a max-pooling layer respectively. The objective of the prediction part is to predict the amount of perceived workload,  i.e. ultimately resulting in a single prediction of the label. The prediction part constitutes a flatten-layer,  with the objective of representing all input into one dimensional shape, followed by a dropout layer, a dense layer, a Rectified Linear (RELU) activation function closed by another dense layer. An overview of the amount of utilized filters / neurons for each layer of each network is provided in Table \ref{table:modelvariations}.

\vspace{7mm}
\begin{figure}[h]
\centering
\includegraphics[scale=0.50]{single_model_architecture}
\caption{Unimodal Network Architectures. Each network constitutes a convolutional part of four convolutional blocks, and a classification part of two fully connected dense layers. }\label{fig:singlearchitecture}
\end{figure}

\newpage
We optimized the hyperparamater for learning rate, dropout rate and the amount of neurons for the first dense layer (\enquote{Dense1} in Figure \ref{fig:singlearchitecture}). This was managed in a total of 50 trials per network.  With respect to network training, a total of 600 epochs have been made through the data for each network specifically. All HPO and training of the unimodal networks was done on a V100 GPU,  offered by the GPU cloud service Google Collab.  Total running time for the unimodal networks was about 96 hours, the most of which was absorbed by HPO.

@@

\subsubsection*{Multimodal Network Architectures} \label{section:multimodal}
The multimodal network architecture was determined by combining the previously characterized unimodal architectures into a single architecture. These separate network parts were joined by a head-network, the architecture of which drew inspiration from aforementioned research by \citeA{han2020classification}. A visual representation of the multimodal network is depicted in Figure \ref{fig:multiarchitecture}. An overview of the amount of utilized filters and neurons is provided in Table \ref{table:modelvariations}.

The feature extraction part of the multimodal network architecture is highly similar to the separate unimodel architectures. The seperate unimodal parts were combined by means of an intermediate fusion strategy, due to its highly flexible nature as compared with alternative fusion strategies (See: Section \ref{sec:related}). The outputs of the three distinct subparts were flattened, followed by a dense layer and RELU activation.  Flattening was necessary such that all inputs were represented into one-dimensional space, after which concatenation rendered possible. The prediction part, also referred to as "Head-Network", consists of three dense layers, alternated with a dropout layer, batch normalization layers and RELU activation layers.

We optimized the hyperparameter for learning rate, dropout rate and the amount of neurons for the first and second dense layers in 18 trials per network.  For training, a total of 250 epochs have been made through the data for each network independently. All training and HPO has been done on a Cloud TPU v2,  offered by the GPU cloud service Google Collab Pro.  Total running time for the multimodal networks was about 7 days, the most of which absorbed by HPO.

\vspace{6mm}
\begin{figure}[h]
\centering
\includegraphics[scale=0.50]{multi_model_architecture}
\caption{Multimodal Network Architecture. The network constitutes a convolutional part of four convolutional blocks for each of distinct network parts. All three distinct parts were flattened and concatenated subsequently. The prediction part of the network constitutes several dense layers.}\label{fig:multiarchitecture}
\end{figure}

\subsection{Network Performance Evaluation}
For each of the four investigated network architectures, the predictions of all 27 individually trained network were aggregated, after which performance was assessed by means of the following three performance metrics.

>>>>

\title{Physiological sensor based mental workload assessment: A multimodal deep learning approach}

@@

\begin{abstract}
The approach of using physiological sensors for the assessment of mental workload has gained in popularity in recent years. We utilized three physiological sensors in predicting mental workload, being electroencephalography (EEG), photoplethysmography (PPG) and galvanic skin response (GSR). Deep learning poses a promising, yet novel, approach towards the modeling of such sensory data. Four deep neural network architectures were created and contrasted in their performance. Three out of four were specified to predict workload based on on a single of the aforementioned physiological sensors, whilst the fourth architecture was a joint neural network combining all physiological signals into a single model. In this study we investigated 1) which physiological sensor individually constitutes the most adequate predictor of mental workload, and 2) whether an approach utilizing multiple sensors simultaneously (i.e. multimodal) is preferable over the approaches utilizing just one. Altogether the EEG-only architecture was found to perform best displaying a mean absolute error of 0.110. We thus conclude that EEG may be considered the most adequate predictor of workload. A mean absolute error of 0.119 was found for the GSR-only architecture. Against expectations, the multimodal approach was outperformed by both aforementioned architectures, displaying a mean absolute error of 0.128. The PPG-only architecture was found to perform worst, displaying a mean absolute error of 0.136. We posed several explanations that help rationalize why the multimodal architecture was found to perform suboptimal, one being an imperfect model architecture, and another being overfitting as resulting from a limited amount of data. Regardless, we demonstrated that the differences in performance across the four network architectures were rather small, i.e. all four neural network architectures performed adequately. 
\end{abstract}

@@

\begin{Large}
{\setlength{\parindent}{0cm}
Methodology and Statistics for the Behavioural, Biomedical and Social Sciences
}
\vspace{2mm}
\newline Utrecht University, the Netherlands
\vspace{18mm}
\newline MSc Thesis Bart-Jan Boverhof (6000142)
\vspace{2mm}
\newline Physiological sensor based prediction of mental workload: A multimodal deep learning approach
\vspace{2mm}
\newline May 2021
\vspace{18mm}
\newline Supervisor:
\vspace{1mm}
\newline Prof. Dr. Ir. Bernard Veldkamp
\vspace{10mm}
\newline Second grader:
\vspace{1mm}
\newline Prof. Dr. Ir René Eijkemans
\vspace{10mm}
\newline Preferred journal of publication: Frontiers in Neuroscience
\vspace{1mm}
\newline Word count: 5603
\end{Large}
\end{titlepage}

@@

A commonly employed method for the assessment of workload in an experimental setting is the well established NASA-Task Load Index questionnaire, hereafter referred to as \enquote{NASA-TLX}. This method embodies several dimensions, jointly inquiring into the amount of perceived workload \cite{hart2006nasa}. Recently, a novel approach has gained in popularity, however. This promising approach constitutes the use of physiological sensors, worn throughout the entire duration of the experiment. The advantage of this approach, is that the development of perceived workload may be monitored throughout the entire experiment, enabling the possibility to precisely monitor the impact of a specific task on a participant's level of workload. Another advantage is that the method may be personalized, which could simply be done by creating and training an individual model for each participant. By doing this, the person-specific character that is inherent to workload is acknowledged, the importance of which was stressed in the opening paragraph of this paper.

With the current research we explore the feasibility of an approach towards workload assessment by means of various physiological sensors, hereafter referred to as \enquote{modalities}. The three utilized modalities are electroencephalography (EEG), photoplethysmography (PPG) and galvanic skin response (GSR). EEG is a technique that monitors electrical activity in the brain, and is often utilized in the assessment of workload \cite{berka2005evaluation, craik2019deep, hogervorst2014combining}. GSR is a technique that measures electrical conductance of the skin, and equally so is a widely adopted approach to workload assessment \cite{nourbakhsh2012using, zhou2015dynamic}. Lastly, also heart-rate is a widely utilized indicator of workload, mostly obtained through PPG \cite{zhang2018evaluating,jimenez2018using}.

In this project we opted for a deep learning approach towards modeling. A range of four different deep neural network architectures, hereafter referred to as \enquote{DNN's} or simply \enquote{networks},  are constructed and contrasted in their performance. Among these are three unimodal DNN architectures, each utilizing only a single modality. Thus,  one architecture solely utilizes the EEG modality,  one solely utilizes the GSR modality and one solely utilizes the PPG modality. The fourth is a multimodal architecture, combining all three aforementioned modalities into a single framework. The main advantage of taking a deep learning approach resides its flexibility: designing a network architecture that combines multiple modalities may be done in a relatively straightforward manner \cite{ramachandram2017deep}.

With this research we 1) seek to explore which individual modality constitutes the most adequate predictor of workload when taking a deep learning approach, and 2) whether a multimodal DNN architecture is preferable over the three simpler (but much less computationally demanding) unimodal network architectures. With regards to the first point, only a single study was found that contrasted several modalities in their ability to predict workload. \citeA{hogervorst2014combining} contrasted multiple modalities, and reported EEG to be the best performing.We therefore carefully hypothesize the EEG modality to be the most adequate predictor of workload. With regards to the second point, we expect the multimodal approach to attain better performance as compared with the unimodal approaches, for predictions are based on richer composite of information. This expectation is indeed in line with previous research, which reported a multimodal approach to perform superiorly \cite{dolmans2020perceived, han2020classification,rastgoo2019automatic, yin2017recognition}. Taking a deep learning approach such as ours may still be considered a novel, yet promising, approach to workload assessment. Due to the novelty of this approach, a deepening of the body of knowledge is necessary, which is something we aim to contribute to with this study.

@@

\subsubsection*{EEG Network Architecture}
An overview of the complete literature on EEG applications with deep learning was presented by \citeA{craik2019deep}, who reported 16\% of all publications to constitute workload assessment. \citeA{tabar2016novel} proposed a ConvNet and stacked auto-encoder (SAE) hybrid approach towards workload classification with EEG. Their proposed network constituted two different parts. Firstly, raw EEG data was fed into a single convolutional layer, with the objective of extracting features. Secondly, the extracted features were fed into the SAE part of the network,  which encompassed a stack of dense-layers designed to ultimately result in a classification of workload.  An accuracy of 90\% was acquired with this architecture \cite{tabar2016novel}. A similar approach was proposed by \citeA{schirrmeister2017deep}, who contrasted different DNN architectures in their ability to adequately decode EEG signals. Several ConvNet architectures were contrasted against the baseline method for EEG decoding, filter bank common spatial pattern (FBCSP). A deep ConvNet, a shallow ConvNet, a deep-shallow hybrid ConvNet and a residual ConvNet were contrasted with the FBCSP baseline method. Both the deep and shallow ConvNets were found to reach at least similar, and in some regards better, classification results as compared with the FBCSP baseline approach. Altogether, a deep ConvNet with four convolutional-max-pooling blocks was found to perform best, exhibiting a classification accuracy of 92.4\% \cite{schirrmeister2017deep}.

\subsubsection*{GSR Network Architecture}
\citeA{sun2019hybrid} explored various deep-learning approaches towards the classification of emotional states with GSR. Despite that this application differs from workload prediction, inspiration may still be drawn from their utilized architecture. Various models were explored, amongst others a support vector machine, a ConvNet and a long-short-term-memory (LSTM) network. In addition to these, a hybrid ConvNet-LSTM approach was explored as well. This hybrid approach constituted four convolutional blocks, followed by a single LSTM layer, and was found to perform best with an accuracy of 74\% \cite{sun2019hybrid}. \citeA{dolmans2020perceived} contrasted several physiological signals in their ability to predict workload, each of which by taking a deep learning approach. One of these modalities constituted GSR, for which a ConvNet-LSTM hybrid network architecture was utilized to predict workload. The performance of this DNN was contrasted with a network consisting solely of fully connected dense layers. Conform with the previously mentioned findings by  \citeA{sun2019hybrid}, the ConvNet-LSTM hybrid architecture was found to perform best, displaying an absolute difference between predicted and true label of 0.180 (scaled on 0-1). This network architecture incorporated two convolutional blocks, followed by two LSTM layers \cite{dolmans2020perceived}.

@@

\subsubsection*{Multimodal Fusion Strategy}  
When creating a DNN that combines multiple modalities into a single architecture,  the information streams that stem from the different subparts of the network (each corresponding to one of the modalities) are required to be combined, i.e. \enquote{fused}, at a given point in the network. This is necessary in order to ultimately end up with a single prediction of workload. Fusion can be done conforming several strategies, out of which three strategies as proposed by \citeA{ramachandram2017deep} are considered.

Firstly, early fusion constitutes an approach that fuses data sources before being fed into the network. Early fusing usually proves to be quite challenging, residing in the fact that data streams stemming from different modalities often differ in their dimensionality and sampling rate. In addition, when taking an early fusion approach, the assumption of conditional independence is made implicitly. This assumption is oftentimes unrealistic in practice, for data stemming from different modalities are expected to be correlated \cite{ramachandram2017deep}. Secondly, late fusion refers to the process of aggregating the decisions of multiple separate networks, each constructed for every modality separately. When the data sources stemming from the various modalities happen to be either correlated, or different in dimensionality, late fusion is oftentimes a more feasible approach as opposed to early fusion \cite{ramachandram2017deep}. Lastly, intermediate fusion is the most widely employed fusion strategy for multimodal networks. Data streams are usually fused by a concatenation layer, joining the outputs of the separately defined network parts of each modality. This results in a single joint deep neural network. Several higher-order layers are usually defined in between the concatenation layer and the ultimate prediction layer. The depth of the fusion, i.e. the specified number of higher-order layers, may be chosen by the researcher, posing intermediate fusion to be the most flexible and therefore the most widely adopted fusion strategy \cite{ramachandram2017deep}.

@@

\subsubsection*{Auxiliary Layers in a DNN Architecture}
At the core of a DNN are the previously mentioned types of layers, such as convolutional, LSTM and dense layers. Such layers are oftentimes accompanied with auxiliary layers. The first type of layers often used in network architectures are batch normalization layers, which have the objective of improving network stability. It is beneficial to incorporate a batch normalization layer subsequent to a convolutional layer, but before feeding into the activation layer \cite{ioffe2015batch}.  For a deep dive into batch normalization, we refer the reader to \citeA{ioffe2015batch}. All previously considered network architectures made use of batch normalization layers \cite{biswas2019cornet, dolmans2020perceived, schirrmeister2017deep, sun2019hybrid, tabar2016novel}.

Pooling layers are commonly employed in ConvNets, usually succeeding a convolutional layer with the purpose of reducing dimensionality. The objective of such layers are to down-sample features into a more compact space, hereby only retaining essential information, and thus omitting redundant information. For a more extensive elaboration on pooling we refer the reader to \citeA{lecun2015deep}. All previously considered network architectures utilized max-pooling layers, usually specified subsequent to the activation layer \cite{biswas2019cornet, dolmans2020perceived, schirrmeister2017deep, sun2019hybrid, tabar2016novel}.

The incorporation of dropout was initially proposed by \citeA{srivastava2014dropout}. Dropout layers drop a randomly-selected number of neurons and their connections throughout the training process. More specifically, within each full training run through the data, i.e. each \enquote{epoch}, a different randomly selected set of neurons are fixed to be inactive. Doing so helps to reduce overfitting, particularly for complex network architectures \cite{srivastava2014dropout}.  Most of the previously considered network architectures made use of dropout layers \cite{biswas2019cornet, dolmans2020perceived, schirrmeister2017deep, sun2019hybrid}.

\subsubsection*{Hyperparameter Optimization}
Hyperparameter optimization (HPO) is a technique that can optimize various DNN hyperparameters, such as for example learning rate and dropout rate. Substantial advancements in DNN performance have been attained by utilizing HPO, especially in the case of ConvNets \cite{bergstra2012random}. In theory, one could optimize the entire DNN architecture, including the amount of neurons/filters in dense/convolutional layers, the number of layers in general, whether to use certain layers, etc. There is, however, a strong relationship between the amount of parameters to be optimized and the computational resources required to do so: optimizing many hyperparameters will inflate computational costs substantially. The Optuna toolbox provides a method for creating a parameter search space, out of which values for the hyperparameters may be sampled \cite{akiba2019optuna}.

@@

\subsubsection*{Experimental Setup \& Data Labeling Strategy}
The experimental setting in which our data was collected is the open-source spaceship video-game Empty Epsilon, in which partakers carried out tasks on a virtual spaceship bridge \cite{daid2016empty}. This experiment was instituted by the Brain Computer Interfaces Testbed\footnote{https://bmslab.utwente.nl/} hosted by the University of Twente and carried out in cooperation with Thales Group Hengelo, the Netherlands\footnote{https://www.thalesgroup.com/en/countries/europe/netherlands}. The experiment constituted 15 tasks, differing in difficulty and hence designed to evoke varying degrees of workload. The total duration of the experiment was roughly three hours.

Respondents were asked to fill out the NASA-TLX questionnaire after each of the 15 task, inquiring into the amount workload perceived during the respective task. Each questionnaire consisted out of six items. The mean of a subset of four of these items were utilized for labeling of the data. Two items were not incorporated into the scale, for they inquired into the perceived physical demands. Since the experiment took place on a computer whilst participants sat behind a desk, we chose to not incorporate such items. Each item, and hence the newly constructed scale, ranges from 0-20, wherein 0 reflects the lowest possible level of workload and 20 reflects the highest possible level of workload.  In order to encourage numerical stability, label scores were normalized to reside in-between 0-1.

@@

\subsubsection*{Data Preprocessing \& Partitioning}
For each participant, the EEG, PPG and GSR data streams were cut into windows of eight seconds, resulting in an average of roughly 351 windows per respondent. 80\% of the windows have been used for training, 10\% for validation within the training process and 10\% for testing final network performance. The partitioning of the windows into training, validation and testing sets was done systematically by selecting every \textit{n-th} window, such that windows over the entire duration of the experiment were represented equally in each partition.  Exactly the same partitions were utilized in the training, validation and testing procedure across the four architectures, in order to ensure a fair comparison.

\subsection{Network Architectures \& Training} \label{section:DNN}
As was shortly elaborated on in the introduction, an advantage of our approach is that an individual network may be trained for every participant separately.  This is helpful, since individuals tend to respond differently towards high workload situations: for some, workload manifests itself mostly through an increase in heart rate detected through PPG, whereas for others, it mainly manifests as an increase in skin conductivity detected through GSR. Since a total of four architectures are contrasted for 27 participants, this resulted in a total of 108 trained networks. Equally so, HPO was conducted for each person individually, resulting in 108 sets of optimized hyperparameters. All networks were constructed in Python, by utilizing the deep-learning toolbox PyTorch \cite{paszke2017automatic}.

\subsubsection*{Unimodal Network Architectures}
All unimodal architectures drew inspiration from aforementioned research \cite{biswas2019cornet, dolmans2020perceived, schirrmeister2017deep, sun2019hybrid}. We opted for a ConvNet architecture for all three unimodal approaches.

The three unimodal DNN architectures are depicted alongside one another in Figure \ref{fig:singlearchitecture}. Each of the unimodal networks may be distinguished by two separate parts, being the convolutional part and the prediction part.  The objective of the convolutional part is to extract features from the data, and consists of four convolutional blocks. Each convolutional block consists of a convolutional layer, a batch normalization layer, the Exponential Linear Unit (ELU) activation function, and a max-pooling layer respectively. The objective of the prediction part is to predict the amount of perceived workload,  i.e. ultimately resulting in a single prediction of the label. The prediction part constitutes a flatten-layer,  with the objective of representing all input into one dimensional shape, followed by a dropout layer, a dense layer, a Rectified Linear (RELU) activation function, closed by another dense layer. An overview of the amount of utilized filters / neurons for each layer of each architecture is provided in Table \ref{table:modelvariations}.

\vspace{7mm}
\begin{figure}[h]
\centering
\includegraphics[scale=0.50]{single_model_architecture}
\caption{Unimodal network architectures. Each network architecture constitutes a convolutional part of four convolutional blocks, and a classification part of two fully connected dense layers. }\label{fig:singlearchitecture}
\end{figure}

\newpage
We optimized the hyperparamaters learning rate, dropout rate and the amount of neurons for the first dense layer (\enquote{Dense1} in Figure \ref{fig:singlearchitecture}). This was managed in a total of 50 trials per network.  With respect to network training, a total of 600 epochs have been made through the data for each network specifically. All HPO and training of the unimodal networks was done on a V100 GPU,  offered by the GPU cloud service Google Collab.  Total running time for the unimodal networks was about 96 hours, the most of which was absorbed by HPO.

@@

\subsubsection*{Multimodal Network Architectures} \label{section:multimodal}
The multimodal network architecture was designed by combining the previously characterized unimodal architectures into a single architecture. These separate network parts were joined by a head-network, the architecture of which drew inspiration from aforementioned research by \citeA{han2020classification}. A visual representation of the multimodal network is depicted in Figure \ref{fig:multiarchitecture}. An overview of the amount of utilized filters and neurons per layer is provided in Table \ref{table:modelvariations}.

The feature extraction part of the multimodal network architecture is highly similar to the separate unimodel architectures. We opted for an intermediate fusion strategy, due to its highly flexible nature as compared with alternative fusion strategies (See: Section \ref{sec:related}). The outputs of the three distinct subparts were flattened, followed by a dense layer and RELU activation funtion. Flattening was necessary such that all inputs were represented into one-dimensional space, after which concatenation becomes possible. The prediction part, also referred to as "head-network", consists of three dense layers, alternated with a dropout layer, batch normalization layers and RELU activation layers.

We optimized the hyperparameters learning rate, dropout rate and the amount of neurons for the first and second dense layers in 18 trials per network.  For training, a total of 250 epochs have been made through the data for each network independently. All training and HPO has been done on a Cloud TPU v2,  offered by the GPU cloud service Google Collab Pro.  Total running time for the multimodal networks was about 120 hours, the most of which absorbed by HPO.

\vspace{6mm}
\begin{figure}[h]
\centering
\includegraphics[scale=0.50]{multi_model_architecture}
\caption{Multimodal network architecture. The network constitutes a convolutional part of four convolutional blocks for each of three subparts. All three distinct parts were flattened and concatenated. The prediction part of the network constitutes several dense layers.}\label{fig:multiarchitecture}
\end{figure}

\subsection{Network Performance Evaluation}
As the reader might recall from Section \ref{section:DNN} we opted for individually tailored DNN's, trained for each respondent separately. Hence, for each of the the four contrasted architectures, the predictions of all personally trained network were aggregated, after which performance was assessed by means of the following three performance metrics.
